 
 
---
title: "Before Starting"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float:
      toc_depth: 3
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```
## About R 

R is an open-source programming language often used for research purposes. You'll find out more about R in the next lesson, Introduction to R. We need to install R, as well as RStudio (software that improves the usability and appearance of R), and many of its associated packages.  


### Installing R

To securely download **R**:

-   Go to <https://cran.r-project.org/>

-   Under 'Download and Install R', select the download option appropriate for your operating system (OS)


-   Download the file appropriate for your OS:

    -   For Windows users, click 'Base'. This will begin downloading the installer for the latest R version for your OS.

    -   For Mac users, select the latest R version`.pkg` file appropriate for your OS version. The installer for R will begin downloading  
   

-   Follow the instructions of the installer


### Installing RStudio:

-   Go to RStudio for desktop <https://www.rstudio.com/products/rstudio/download/#download>

-   Select the install file appropriate for your OS and follow the installer instructions

For **Windows -** Install **Rtools** (an R tool needed to build R packages with C/C++/Fortran code from source) by visiting this page: <https://cran.r-project.org/bin/windows/Rtools/rtools40.html>


### Updating R

At some point, running old versions of software adds unnecessary difficulty. R had a major version change in April 2020, with the release of 4.0.0. It is a good idea to be on the current major version, meaning 4.3.0 or higher as of 2024. Each major version is followed by several years of smaller releases (minor and patch releases) that are indicated by the numbers following '4.' (ex. 4.3.2, 4.3.1, etc.,). You can be more relaxed about upgrading minor versions (such as upgrading between 4.3.1 and 4.3.2), but you still want to stay reasonably current. A consequence of running 'old' versions of R are that functions may not operate correctly, creating unnecessary (and sometimes undetectable!) errors. 

You can check your current R version by running the code below in your RStudio console pane:

```{r}
# Here we use function 'R.Version()' to receive information about the R version we are currently using: 
R.Version()
#> [1] "R version 4.3.2 (2023-10-31)"
```


### Updating RStudio

To see if this is the most recent version of R, go to https://cran.r-project.org/. Alternatively, you can go to `Help`, the right-most option in the R toolbar. Select 'Check for Updates' to receive instructions and assistance installing the latest version of R. 

Warning: Running a different major version of R (such as R 3.6.3) will almost definitely cause *significant* difficulties. It is also deeply beneficial to use an R version with the second digit matching the most current version (ie., using 4.3.2 is preferable to using 4.2.2). Upgrading R when the third digit is different (indicating a minor upgrade) can be considered optional (i.e., upgrading between R 4.3.2 and 4.3.1 will have minor differences). 


## About R Packages

Packages are extra collections of code that add new capabilities to R. You can think of them as analogous to DLC in video games. There are hundreds of R packages available, and since R is an open software, anyone can create an R package. 

There are three essential tasks that will enable the use of a package in R. 
-   **Installation:** Packages must be installed, typically with `install.package()`. You only need to do this *once* per major R version (4.2.0 vs. 4.3.0).

```{r, message=FALSE, warning=FALSE}
# Here we install the package 'dplyr', a package for manipulating data.
install.package("dplyr")
```


-   **Loading:** Each time you open an R session, you need to load the package before you can use it. You can do this with the `library()` function. 

```{r, message=FALSE, warning=FALSE}
# Here we load dplyr.
library("dplyr")
```


-   **Updating:** Just like R, packages will receive upgrades periodically. It is important to keep up-to-date, because discrepancies between packages and the version of R they were written for can cause problems. You can use `update.packages()` to update any packages that are out of date. Alternatively, you can go to 'Packages' and select 'Update'. This will show you a preview of all available updates. 

```{r, message=FALSE, warning=FALSE}
# Update any packages with more recent versions. 
update.packages()
```


### Installing R packages

How a package is installed depends on where it comes from. Packages hosted by the Comprehensive R Archive Network (CRAN) use a standard installation command (`install.packages()`). Many non-CRAN packages can also be installed in this fashion. The rest will require you to find instructions and may depend on other packages, such as BiocManager. 


#### CRAN Packages

Roughly 30 packages will be necessary in this series of lessons. Instead of loading each package individually (which can become tedious), we can create a list of the packages we want, and then pass that list to `library()`. 

In the next chunk of code, we describe the purpose of each package. This is purely educational - you do not need to do this in your own R scripts. 

**Tip:** You only need to install packages once, barring a major update. If a new installation is needed, R will tell you. 

```{r , message=FALSE, warning=FALSE}

# Here we create a list of desired packages, expressed as character strings. (You'll learn more about different data structures in the next lesson).

list.of.packages <-
  c("devtools", # A package for managing (installing, loading, updating) other packages 
    "conflicted", # A package for managing 'conflicts' within packages. A conflict is when two packages have functions with the same name but a different purpose
    "here", # A package that improves reproducibility across different operating systems/users by defining the 'work environment', or base of operations. 
    
    # There are also packages for managing data files:
    "readr", # A package for handling common data files like `.csv`
    "readxl", # A package for handling Excel files like `.xlsx`
    
    # Here are packages that help manipulate (or 'wrangle') data structures:
    "dplyr", # For data manipulation, such as inspecting individual variables
    "tidyr", # For 'tidying' data, such as converting the data's structure or making it more readable
    "reshape2", # Another package with additional tools for manipulating your data
    "purrr", # A package that provides tools for working with functions and vectors
    "openxlsx", # Allows for reading and writing Excel files
    
    # Packages for plotting: 
    "ggplot2", # A popular R package with tools for creating high-level plots
    "plotly", # Another advanced visualization package
    "ggsci", # A package for applying color palettes to visuals
    "ggthemes", # A package that adds more options for visual design
    "patchwork", # A package that allows users to combine plots
    "stringr", # A package that can improve the readability of visuals by providing tools for string and factor manipulation (i.e., 'trt' can become 'treatment' for the purposes of creating an informative figure)

    # Packages for analysis & modelling:
    "minpack.lm", # For fitting non-linear models
    "deSolve", # For solving initial value problems of differential equations
    "psych", # A package with tools for evaluating psychiatric data. We use this as an educational example
    "car", # For more analytical/modelling functions and tools
    "emmeans", # For modelling functions
    "agricolae", # A source of diverse functions and data sets in agronomy 
    
    # Packages for visualizing outputs of statistical analysis: 
    "rcompanion", # A useful tool for summarizing model outputs and other modeling tasks
    "ggResidpanel", # For creating and managing diagnostic plots

    # Packages for parallel computation: 
    # (using multiple cores to execute R code)
    "pbapply", # A package that visualizes computational progress
    "parallel", # A package for managing concurrent computational processes 

    # Packages for text formatting/report creation:
    "rmarkdown", # A package for creating and formatting files (such as htmls) in R
    "gt", # A package for customizing html table view
    "kableExtra", # Another package for html table customization
    
    # Advanced visualizations/App building:
    "shiny", # Package for visualization 
    "shinyscreenshot", # For capturing screenshots
    "shinydashboard" # Additional shiny tools
  )

# Now that we have a list of packages, we can load them
install.packages(list.of.packages)
```

We have installed our desired packages by passing a list to `install.packages()`. This is must easier than using `install.packages()` for each of our ~30 desired packages. 


##### Troubleshooting 
Installation should go smoothly if your R version is up to date, but there can be occasional problems. Below, we have shown how to deal with several of the more common errors that can occur during package installation. 

If there are problems installing some package, sometimes directing R to download packages from different sources (repositories) can help the issue. CRAN has [multiple repositories](https://cran.r-project.org/mirrors.html) that get different amounts of traffic at different times. Repositories can also slow down because of failures or maintenance. You can direct R to a another, less busy repository with the `repos` argument. 
```{r}
# For example, we can run this line where we direct R to install package ROCR from a different CRAN repository. Just un-comment the command below (remove the #).

install.packages("ROCR", repos = c(CRAN="https://cran.r-project.org/"))

```

Assuming you have listed all of your desired packages in object `list.of.packages`, you can check that all packages are installed with the following command: 

```{r}
list.of.packages %in% installed.packages()
```

If you see some `FALSE` values there, then you can check which packages are missing with the following line of code. 

```{r}
list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]

```


#### Loading R packages

Once your desired packages are installed, you'll need to load them in to R. Packages will need to be loaded each time you open R before they can be used. Just like with installation, we can save time by passing a list of packages to the relevant command, 'library()'. 

```{r, message=FALSE, warning=FALSE}

``` 


#### Updating R packages


##### Windows

It's good to update your installed packages often. You might need to restart R during this process as well. With Windows OS, this is straightforward. 

```{r}
update.packages(ask = FALSE, checkBuilt = TRUE)
```

Additionally, R will often alert you to package updates with feedback in the RStudio console. 


##### Other OS

Updating packages with non-Windows OS can be more complicated. A lot of packages that currently exist were built under the old major R version (R 3.0.0 through 3.6.9). If you have any packages that were downloaded when you were using any R version below R 4.0.0, it is safer to re-install these packages. 

You can achieve this with the following commands. However, this will not work for packages installed with `devtools::install_github()`. 

```{r}
# Get all currently installed packages
packs = as.data.frame(installed.packages(.libPaths()[1]), stringsAsFactors = F)

## Now re-install these packages using install.packages():
install.packages(packs$Package)
```


### Non-CRAN Packages

There are packages which are not published in public repositories. These are installed directly from some public repository, such as GitHub.

```{r}
.rs.restartR()
Sys.sleep(2)
remotes::install_github("bbc/bbplot",dependencies = TRUE)
```

If you run into problems... Installing packages that are not on CRAN can be a pain, so there are a few general notes on what to do.\
If you are Windows user it is recommended that you make sure you have the [Rtools](https://cran.r-project.org/bin/windows/Rtools/) installed. If you get an error like this while updating your packages:

```{r error_noncran_pck, eval=FALSE}
Error: (converted from warning) cannot remove prior installation of package ‘name_of_the_package’
```

You might need to update some packages manually to match the *source* version. So, just replace the `name_of_the_package` to match the name of the package you are missing/need to update.

```{r}
install.packages("name of the package", type = "source")
```

After that, try run the code snippet for the installation of package from the beginning of this subsection.

From here, you can try the trouble-shooting methods we discussed above. If these methods don't work, you can often find package-specific installation instructions with the package's documentation online. For instance, entering 'r install KrigR' in a browser search bar will bring you to the GitHub repository where non-CRAN package 'KrigR' is hosted. Reading the README.md file shows specific installation instructions. From there, you can copy/paste the installation instructions to your RStudio console.

For instance, running these lines in your RStudio console will install and load package KrigR:
```{r}
# Un-comment the next three lines (remove the #) to install KrigR
# Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS="true")
# devtools::install_github("https://github.com/ErikKusch/KrigR")
# library(KrigR)
```


### Install our package with apps

We will need certain packages for this tutorial. Run following lines of code to install them:

```{r eval = FALSE}
if(!"devtools" %in% installed.packages ()) install.packages("devtools")
devtools::install_github("mladencucak/AppTest")
AppTest::GrowModAPP()
AppTest::HLIRApp()
```
